---
title: "Visualizing Census data"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    toc-depth: 4
runtime: shiny_prerendered
description: >
  Chapter 4 of the book [Analyzing US Census Data](https://walker-data.com/census-r/the-united-states-census-and-the-r-programming-language.html).
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(tidycensus)

tutorial_options(
  exercise.timelimit = 60,
  # A simple checker function that just returns the message in the check chunk
  exercise.checker = function(check_code, ...) {
    list(
      message = eval(parse(text = check_code)),
      correct = logical(0),
      type = "info",
      location = "append"
    )
  }
)
knitr::opts_chunk$set(error = TRUE)
```

## Introduction

In this tutorial, we learn to:

* visualize Census data using **ggplot2** and other plotting packages.

This tutorial follow [Analyzing US Census Data](https://walker-data.com/census-r/the-united-states-census-and-the-r-programming-language.html) Chapter 4 and [_R for Data Science_](http://r4ds.had.co.nz/).

## Basic visualization

- Let's start with the data on median household income and median age by county in the state of California from the 2016-2020 ACS.
```{r}
library(tidycensus)

ca_wide <- get_acs(
  geography = "county",
  state = "CA",
  variables = c(medinc = "B19013_001",
                medage = "B01002_001"),
  output = "wide",
  year = 2020
)

ca_wide
```

- Our first **ggplot2** figure is a **histogram** of median income:
```{r}
library(tidyverse)
options(scipen = 999)

ggplot(ca_wide, aes(x = medincE)) + 
  geom_histogram()
```

    - `options(scipen = 999)` instructs R to avoid using scientific notation
    
    -  By default, `geom_histogram` uses 30 bins

```{r}
library(tidyverse)
options(scipen = 999)

ggplot(ca_wide, aes(x = medincE)) + 
  geom_histogram(bins = 15)
```    

- Boxplot:
```{r}
ggplot(ca_wide, aes(y = medincE)) + 
  geom_boxplot()
```

- Relationship between two variables by scatterplot:
```{r}
ggplot(ca_wide, aes(x = medageE, y = medincE)) + 
  geom_point()
```
Overlay with a linear model fit:
```{r}
ggplot(ca_wide, aes(x = medageE, y = medincE)) + 
  geom_point() + 
  geom_smooth(method = "lm")
```
Overlay with a non-linear model fit:
```{r}
ggplot(ca_wide, aes(x = medageE, y = medincE)) + 
  geom_point() + 
  geom_smooth(method = "loess")
```

## Customizing ggplot2 visualization

- Suppose we want to illustrate the percent of commuters that take public transportation to work for the largest metropolitan areas in the United States, using data from the 2019 1-year ACS.
```{r}
library(tidycensus)
library(tidyverse)

metros <-  get_acs(
  geography = "cbsa",
  variables = "DP03_0021P",
  summary_var = "B01003_001",
  survey = "acs1",
  year = 2019
) %>%
  slice_max(summary_est, n = 20)

metros
```

- Bar chart:
```{r}
ggplot(metros, aes(x = NAME, y = estimate)) + 
  geom_col()
```

- Shorten metro names, reorder by the percentage (`estimat`), rotate the bar chart, and more meaningful title and axis labels:
```{r}
metros %>%
  mutate(NAME = str_remove(NAME, "-.*$")) %>%
  mutate(NAME = str_remove(NAME, ",.*$")) %>%
  ggplot(aes(y = reorder(NAME, estimate), x = estimate)) + 
  geom_col() + 
  theme_minimal() + 
  labs(title = "Public transit commute share", 
       subtitle = "2019 1-year ACS estimates", 
       y = "", 
       x = "ACS estimate", 
       caption = "Source: ACS Data Profile variable DP03_0021P via the tidycensus R package") 
```

- More styles:
```{r}
library(scales)

metros %>%
  mutate(NAME = str_remove(NAME, "-.*$")) %>%
  mutate(NAME = str_remove(NAME, ",.*$")) %>%
  ggplot(aes(y = reorder(NAME, estimate), x = estimate)) + 
  geom_col(color = "navy", fill = "navy", 
           alpha = 0.5, width = 0.85) +  
  theme_minimal(base_size = 12, base_family = "Verdana") + 
  scale_x_continuous(labels = label_percent(scale = 1)) + 
  labs(title = "Public transit commute share", 
       subtitle = "2019 1-year ACS estimates", 
       y = "", 
       x = "ACS estimate", 
       caption = "Source: ACS Data Profile variable DP03_0021P via the tidycensus R package") 
```

## Visualizing margins of error

- Let's visualize the median household incomes of counties in the state of California from the 2016-2020 ACS.

- Counties and population sizes:
```{r}
ca <- get_decennial(
  state = "CA",
  geography = "county",
  variables = c(totalpop = "P1_001N"),
  year = 2020
) %>%
  arrange(desc(value))

ca
```

- Median household incomes:
```{r}
ca_income <- get_acs(
  state = "CA",
  geography = "county",
  variables = c(hhincome = "B19013_001"),
  year = 2020
) %>%
  mutate(NAME = str_remove(NAME, " County, California"))

ca_income
```

- Basic plot:
```{r}
ggplot(ca_income, aes(x = estimate, y = reorder(NAME, estimate))) + 
  geom_point(size = 1.5, color = "darkgreen") + 
  labs(title = "Median household income", 
       subtitle = "Counties in California", 
       x = "", 
       y = "ACS estimate") + 
  theme_minimal(base_size = 9) + 
  scale_x_continuous(labels = label_dollar())
```

- Add error bar (from margin of errors):
```{r}
ggplot(ca_income, aes(x = estimate, y = reorder(NAME, estimate))) + 
  geom_errorbarh(aes(xmin = estimate - moe, xmax = estimate + moe)) + 
  geom_point(size = 1.5, color = "darkgreen") + 
  theme_minimal(base_size = 8) + 
  labs(title = "Median household income", 
       subtitle = "Counties in California", 
       x = "2016-2020 ACS estimate", 
       y = "") + 
  scale_x_continuous(labels = label_dollar())
```
What do we observe?
```{r}
ca %>% arrange(value)
```

## Visualize ACS estimates over time

- Let's obtain 1-year ACS data from 2005 through 2019 on median home value for Deschutes County, Oregon, home to the city of Bend and large numbers of in-migrants in recent years from the Bay Area in California. 
```{r}
years <- 2005:2019
names(years) <- years

deschutes_value <- map_dfr(years, ~{
  get_acs(
    geography = "county",
    variables = "B25077_001",
    state = "OR",
    county = "Deschutes",
    year = .x,
    survey = "acs1"
  )
}, .id = "year")

deschutes_value
```

- A stylized plot:
```{r}
ggplot(deschutes_value, aes(x = year, y = estimate, group = 1)) + 
  geom_ribbon(aes(ymax = estimate + moe, ymin = estimate - moe), 
              fill = "navy",
              alpha = 0.4) + 
  geom_line(color = "navy") + 
  geom_point(color = "navy", size = 2) + 
  theme_minimal(base_size = 12) + 
  scale_y_continuous(labels = label_dollar(scale = .001, suffix = "k")) + 
  labs(title = "Median home value in Deschutes County, OR",
       x = "Year",
       y = "ACS estimate",
       caption = "Shaded area represents margin of error around the ACS estimate")
```

## Exploring age and sex structure with population pyramids

- We use data from the Population Estimates API for the state of California:
```{r}
ca <- get_estimates(
  geography = "state",
  state = "CA",
  product = "characteristics",
  breakdown = c("SEX", "AGEGROUP"),
  breakdown_labels = TRUE,
  year = 2019
)

ca
```

- Remove rows for `Both sexes` and flip values of `Male` to negative.
```{r}
ca_filtered <- filter(ca, str_detect(AGEGROUP, "^Age"), 
                  SEX != "Both sexes") %>%
  mutate(value = ifelse(SEX == "Male", -value, value))

ca_filtered
```

- Pyramid plot:
```{r}
ggplot(ca_filtered, aes(x = value, y = AGEGROUP, fill = SEX)) + 
  geom_col()
```

- A more stylized pyramid:
```{r}
ca_pyramid <- ggplot(ca_filtered, 
                       aes(x = value, 
                           y = AGEGROUP, 
                           fill = SEX)) + 
  geom_col(width = 0.95, alpha = 0.75) + 
  theme_minimal(base_family = "Verdana", 
                base_size = 12) + 
  scale_x_continuous(
    labels = ~ number_format(scale = .001, suffix = "k")(abs(.x)),
    limits = 1000000 * c(-1.75, 1.75)
  ) + 
  scale_y_discrete(labels = ~ str_remove_all(.x, "Age\\s|\\syears")) + 
  scale_fill_manual(values = c("darkred", "navy")) + 
  labs(x = "", 
       y = "2019 Census Bureau population estimate", 
       title = "Population structure in California", 
       fill = "", 
       caption = "Data source: US Census Bureau population estimates & tidycensus R package")

ca_pyramid
```

- With aid of the **geofacet** package, we can visualize pyramids of multiple states:
```{r}
library(geofacet)

us_pyramid_data <- get_estimates(
  geography = "state",
  product = "characteristics",
  breakdown = c("SEX", "AGEGROUP"),
  breakdown_labels = TRUE,
  year = 2019
) %>%
  filter(str_detect(AGEGROUP, "^Age"),
         SEX != "Both sexes") %>%
  group_by(NAME) %>%
  mutate(prop = value / sum(value, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(prop = ifelse(SEX == "Male", -prop, prop))

ggplot(us_pyramid_data, aes(x = prop, y = AGEGROUP, fill = SEX)) + 
  geom_col(width = 1) + 
  theme_minimal() + 
  scale_fill_manual(values = c("darkred", "navy")) + 
  facet_geo(~NAME, grid = "us_state_with_DC_PR_grid2",
            label = "code") + 
  theme(axis.text = element_blank(),
        strip.text.x = element_text(size = 8)) + 
  labs(x = "", 
       y = "", 
       title = "Population structure by age and sex", 
       fill = "", 
       caption = "Data source: US Census Bureau population estimates & tidycensus R package")
```

## Visualizing group-wise comparison

TODO

## Interactive visualization with plotly

```{r}
library(plotly)

ggplotly(ca_pyramid)
```

## Exercises

- Choose a different variable in the ACS and/or a different location and create a margin of error visualization of your own.

```{r moe, exercise = TRUE}

```

- Modify the population pyramid code to create a different, customized population pyramid. You can choose a different location (state or county), different colors/plot design, or some combination!

```{r pyramid, exercise = TRUE}

```
